---
title: anndataR
format: gfm
---

:::{.content-hidden}
Rendered using:
```
quarto render README.qmd; sed -i 's#&lt;#<#g;s#&gt;#>#g' README.md
```
:::

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/anndataR)](https://CRAN.R-project.org/package=anndataR)
<!-- badges: end -->


`{anndataR}` aims to make the AnnData format a first-class citizen in the R ecosystem, and to make it easy to work with AnnData files in R, either directly
or by converting it to a SingleCellExperiment or Seurat object.

Feature list:

* Provide an `R6` class to work with AnnData objects in R (either in-memory or on-disk).
* Read/write `*.h5ad` files natively
* Convert to/from `SingleCellExperiment` objects
* Convert to/from `Seurat` objects

:::{.callout-warning}
This package is still in the experimental stage, and may not work as expected. You can find the status of development of anndataR on the [feature tracking page](https://anndatar.data-intuitive.com/articles/design.html#feature-tracking) of the website.Please report any issues you encounter.
:::

## Installation

You can install the development version of `{anndataR}` like so:

``` r
devtools::install_github("scverse/anndataR")
```

You might need to install suggested dependencies manually, depending on the task you want to perform.

* To read/write *.h5ad files, you need to install [hdf5r](https://cran.r-project.org/package=hdf5r):  
  `BiocManager::install("hdf5r")`
* To convert to/from `SingleCellExperiment` objects, you need to install [SingleCellExperiment](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html):  
  `BiocManager::install("SingleCellExperiment")`
* To convert to/from `Seurat` objects, you need to install [SeuratObject](https://cran.r-project.org/package=SeuratObject):  
  `install.packages("SeuratObject")`

If you're feeling adventurous, you can install all suggested dependencies at once:

``` r
devtools::install_github("scverse/anndataR", dependencies = TRUE)
```

## Example

Here's a quick example of how to use `{anndataR}`. First, we download an h5ad file.

```{r warning=FALSE}
library(anndataR)

h5ad_path <- system.file("extdata", "example.h5ad", package = "anndataR")
```

Read an h5ad file in memory:

```{r warning=FALSE}
adata <- read_h5ad(h5ad_path)
```

Read an h5ad file on disk:

```{r warning=FALSE}
adata <- read_h5ad(h5ad_path, to = "HDF5AnnData")
```

View structure:
```{r}
adata
```

Access AnnData slots:
```{r}
dim(adata$X)
adata$obs[1:5, 1:6]
adata$var[1:5, 1:6]
```

## Interoperability

Convert the AnnData object to a SingleCellExperiment object:
```{r}
sce <- adata$to_SingleCellExperiment()
sce
```

```{r include=FALSE}
# load SeuratObject first to avoid warnings related to seurat itsef
library(SeuratObject)
```

Convert the AnnData object to a Seurat object:
```{r}
obj <- adata$to_Seurat()
obj
```

## Manually create an object

```{r}
adata <- AnnData(
  X = matrix(rnorm(100), nrow = 10),
  obs = data.frame(
    cell_type = factor(rep(c("A", "B"), each = 5))
  ),
  var = data.frame(
    gene_name = paste0("gene_", 1:10)
  )
)

adata
```
