---
title: Software design
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Software design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
```


`{anndataR}` is designed to offer the combined functionality of the following packages:

* [theislab/zellkonverter](https://github.com/theislab/zellkonverter): Convert AnnData files to/from `SingleCellExperiment` objects.
* [mtmorgan/h5ad](https://github.com/mtmorgan/h5ad/): Read/write `*.h5ad` files natively using `rhdf5`.
* [dynverse/anndata](https://github.com/dynverse/anndata): An R implementation of the AnnData data structures, uses `reticulate` to read/write `*.h5ad` files.

Ideally, this package will be a complete replacement for all of these packages, and will be the go-to package for working with AnnData files in R.

## Desired feature list

* Provide an `R6` class to work with AnnData objects in R (either in-memory or on-disk).
* Read/write `*.h5ad` files natively
* Convert to/from `SingleCellExperiment` objects
* Convert to/from `Seurat` objects

## Class diagram

Here is a diagram of the main R6 classes provided by the package:

![](diagrams/class_diagram.svg)

Notation:

  - `X: Matrix` - variable `X` is of type `Matrix`
  - `*X: Matrix` - variable `X` is abstract
  - `as_SingleCellExperiment(): SingleCellExperiment` - function `as_SingleCellExperiment` returns object of type `SingleCellExperiment`
  - `*as_SingleCellExperiment()` - function `as_SingleCellExperiment` is abstract

## Session info

```{r}
sessionInfo()
```
