---
title: "Python Integration with anndataR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Python Integration with anndataR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

anndataR works with Python AnnData objects through reticulate.
You can load Python objects, apply Python functions to them, and
convert to Seurat or SingleCellExperiment objects without manual conversion steps.

## Basic Integration with Scanpy

Install required Python packages if needed:

```{r python_setup, eval=FALSE}
reticulate::py_install("scanpy")
```

```{r load_libraries}
library(anndataR)
library(reticulate)
sc <- import("scanpy")
```

Load a dataset directly from scanpy:

```{r load_python_data}
adata <- sc$datasets$pbmc3k_processed()
print(adata)
```

Apply scanpy functions directly:

```{r apply_python_functions}
sc$pp$filter_cells(adata, min_genes = 200L)
sc$pp$normalize_total(adata, target_sum = 1e4)
sc$pp$log1p(adata)
```

Convert to Seurat object:

```{r convert_to_seurat}
seurat_obj <- adata$as_Seurat()
print(seurat_obj)
```

Convert to SingleCellExperiment object:

```{r convert_to_sce}
sce_obj <- adata$as_SingleCellExperiment()
print(sce_obj)
```

## Multi-modal Data with MuData

Install required Python packages if needed:

```{r mudata_setup, eval=FALSE}
reticulate::py_install("mudata")
```

```{r load_mudata}
md <- import("mudata")
```

Load a MuData object from file:

```{r load_mudata_example}
url <- "https://github.com/gtca/h5xx-datasets/raw/b1177ac8877c89d8bb355b072164384b4e9cc81d/datasets/minipbcite.h5mu"
path <- tempfile(fileext = ".h5mu")
download.file(url, path)
mdata <- md$read_h5mu(path)
```

Access individual modalities and convert them:

```{r access_modalities}
rna_mod <- mdata$mod[["rna"]]

rna_seurat <- rna_mod$as_Seurat()
print(rna_seurat)

rna_sce <- rna_mod$as_SingleCellExperiment()
print(rna_sce)
```

