---
title: Read/write H5AD files
package: anndataR
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Read/write H5AD files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates how to read and write `.h5ad` files using the **{anndataR}** package.

Check out `?anndataR` for a full list of the functions provided by this package.

## Example data

A great place for finding single-cell datasets as `.h5ad` files is the [CELLxGENE](https://cellxgene.cziscience.com/datasets) website. 

We will use an example file included in the package for demonstration.

```{r}
library(anndataR)

h5ad_path <- system.file("extdata", "example.h5ad", package = "anndataR")
```

## Reading in memory

To read an h5ad file into memory, use the `read_h5ad` function. By default, the data will be read entirely into memory:

```{r}
adata <- read_h5ad(h5ad_path)
```

This reads the entire `.h5ad` file into memory as an `AnnData` object. You can then inspect its structure:

```{r}
adata
```

## Reading backed by HDF5

For large datasets that do not fit into memory, you can read the h5ad file in a "backed" mode. This means that the data remains on disk, and only parts that are actively being used are loaded into memory. 

To do this, set the `to` parameter in the `read_h5ad` to `HDF5AnnData`:

```{r}
adata <- read_h5ad(h5ad_path, to = "HDF5AnnData")
```
The structure of the object will look similar to the in-memory representation, but the data is stored on disk.

```{r}
adata
```

Note that any changes made to the object will be reflected in the `.h5ad` file!

## Writing h5ad files

You can write an `AnnData` object to an h5ad file using the `write_h5ad` function:

``` r
# Create a temporary file for demonstration
temp_h5ad <- tempfile(fileext = ".h5ad")

adata$write_h5ad(temp_h5ad)
```

## Accessing AnnData slots

The `AnnData` object is a list-like object containing various slots. Here's how you can access some of them:

```{r}
dim(adata$X)
adata$obs[1:5, 1:6]
adata$var[1:5, 1:6]
```

You can also access other slots like `layers`, `uns`, `obsm`, `varm`, and `obsp` in a similar way.

## Session info

```{r}
sessionInfo()
```
