---
title: Read/write Seurat objects using anndataR
package: anndataR
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Read/write Seurat objects using anndataR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
```

This vignette demonstrates how to read and write `Seurat` objects using the **{anndataR}** package, leveraging the interoperability between `Seurat` and the `AnnData` format.

Check out `?anndataR` for a full list of the functions provided by this package.

## Introduction

Seurat is a widely used toolkit for single-cell analysis in R.
 **{anndataR}** enables conversion between `Seurat` objects and `AnnData` objects,
allowing you to leverage the strengths of both the scverse and Seurat ecosystems.

## Prerequisites

Before you begin, make sure you have both Seurat and **{anndataR}** installed. You can install them using the following code:

```r
if (!requireNamespace("pak", quietly = TRUE)) {
    install.packages("pak")
}
pak::pak("Seurat")
pak::pak("scverse/anndataR")
```

## Converting an AnnData Object to a Seurat Object

Using an example `.h5ad` file included in the package, we will demonstrate how to read an `.h5ad` file and convert it to a `Seurat` object.

```{r prep_file}
library(anndataR)
library(Seurat)

h5ad_file <- system.file("extdata", "example.h5ad", package = "anndataR")
```

Read the `.h5ad` file and convert it to a `Seurat` object:

```{r read_data}
seurat_obj <- read_h5ad(h5ad_file, as = "Seurat")
seurat_obj
```

This is equivalent to reading in the `.h5ad` file and explicitly converting.

```{r convert_seurat}
adata <- read_h5ad(h5ad_file)
seurat_obj <- adata$as_Seurat()
seurat_obj
```

Note that there is no one-to-one mapping possible between the AnnData and SeuratObject data structures,
so some information might be lost during conversion. It is recommended to carefully inspect the converted object
to ensure that all necessary information has been transferred.

### Customizing the conversion

You can customize the conversion process by providing specific mappings for each slot in the Seurat object.
Each of the mapping arguments can be provided with one of the following:
- `TRUE`: all items in the slot will be copied using the default mapping
- `FALSE`: the slot will not be copied
- a (named) character vector: the names are the names of the slot in the Seurat object, the values are the names of the slot in the AnnData object.

See `?as_Seurat` for more details on how to customize the conversion process. For instance:

```{r customize_seurat_conversion}
seurat_obj <- adata$as_Seurat(
  layers_mapping = c("counts", "dense_counts"),
  object_metadata_mapping = c(metadata1 = "Int", metadata2 = "Float"),
  assay_metadata_mapping = FALSE,
  reduction_mapping = list(
    pca = c(key = "PC_", embeddings = "X_pca", loadings = "PCs"),
    umap = c(key = "UMAP_", embeddings = "X_umap")
  ),
  graph_mapping = TRUE,
  misc_mapping = c(misc1 = "Bool", misc2 = "IntScalar")
)
seurat_obj
```

## Convert a Seurat Object to an AnnData Object

The reverse conversion is also possible, allowing you to convert the `Seurat` object back to an `AnnData` object, or to just write out the `Seurat` object as an `.h5ad` file.

```{r write_seurat}
write_h5ad(seurat_obj, tempfile(fileext = ".h5ad"))
```

This is equivalent to converting the `Seurat` object to an `AnnData` object and then writing it out:
```{r convert_to_anndata}
adata <- as_AnnData(seurat_obj)
write_h5ad(adata, tempfile(fileext = ".h5ad"))
```

You can again customize the conversion process by providing specific mappings for each slot in the `AnnData` object.
For more details, see `?as_AnnData`.

Here's an example:
```{r customize_anndata_conversion}
adata <- as_AnnData(
  seurat_obj,
  assay_name = "RNA",
  x_mapping = "counts",
  layers_mapping = c("dense_counts"),
  obs_mapping = c(RNA_count = "nCount_RNA", metadata1 = "metadata1"),
  var_mapping = FALSE,
  obsm_mapping = list(X_pca = "pca", X_umap = "umap"),
  obsp_mapping = TRUE,
  uns_mapping = c("misc1", "misc2")
)
adata
```

## Session info

```{r}
sessionInfo()
```
