---
title: Read/write SingleCellExperiment objects using anndataR
package: anndataR
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Read/write SingleCellExperiment objects using anndataR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
```

This vignette demonstrates how to read and write `SingleCellExperiment` objects using the **{anndataR}** package, leveraging the interoperability between `SingleCellExperiment` and the `AnnData` format.

## Introduction

`SingleCellExperiment` is a widely used class for storing single-cell data in R, especially within the **Bioconductor** ecosystem.
**{anndataR}** enables conversion between `SingleCellExperiment` objects and `AnnData` objects, allowing you to leverage the strengths of both the **scverse** and **Bioconductor** ecosystems.

## Prerequisites

This vignette requires **{SingleCellExperiment}** in addition to **{anndataR}**.
You can install them using the following code:

```r
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("SingleCellExperiment")
```

## Converting an `AnnData` object to a `SingleCellExperiment` object

Using an example `.h5ad` file included in the package, we will demonstrate how to read an `.h5ad` file and convert it to a `SingleCellExperiment` object.

```{r prep_h5ad_file}
library(anndataR)
library(SingleCellExperiment)

h5ad_file <- system.file("extdata", "example.h5ad", package = "anndataR")
```

Read the `.h5ad` file as a `SingleCellExperiment` object:

```{r read_data}
sce_obj <- read_h5ad(h5ad_file, as = "SingleCellExperiment")
sce_obj
```

This is equivalent to reading in the `.h5ad` file and explicitly converting.

```{r read_h5ad}
adata <- read_h5ad(h5ad_file)
sce <- adata$as_SingleCellExperiment()
sce
```

Note that there is no one-to-one mapping possible between the `AnnData` and `SingleCellExperiment` data structures, so some information might be lost during conversion.
It is recommended to carefully inspect the converted object to ensure that all necessary information has been transferred.

### Customizing the conversion

You can customize the conversion process by providing specific mappings for each slot in the `SingleCellExperiment` object.

Each of the mapping arguments can be provided with one of the following:
- `TRUE`: all items in the slot will be copied using the default mapping
- `FALSE`: the slot will not be copied
- A (named) character vector: the names are the names of the slot in the `SingleCellExperiment` object, the values are the names of the slot in the `AnnData` object.

See `?as_SingleCellExperiment` for more details on how to customize the conversion process. For instance:

```{r ex_mapping, eval=FALSE}
sce_obj <- adata$as_SingleCellExperiment(
  x_mapping = "counts",
  assays_mapping = c("csc_counts"),
  colData_mapping = c("Int", "IntNA"),
  rowData_mapping = c(rowdata1 = "String", rowdata2 = "total_counts"),
  reducedDims_mapping = list(
    "pca" = c(sampleFactors = "X_pca", featureLoadings = "PCs"),
    "umap" = c(sampleFactors = "X_umap")
  ),
  colPairs_mapping = TRUE,
  rowPairs_mapping = FALSE,
  metadata_mapping = c(value1 = "Bool", value2 = "IntScalar")
)
sce_obj
```

The mapping arguments can also be passed directly to `read_h5ad()`.

## Convert a `SingleCellExperiment` object to an `AnnData` object

The reverse conversion is also possible, allowing you to convert a `SingleCellExperiment` object back to an `AnnData` object, or to just write out the `SingleCellExperiment` object as an `.h5ad` file.

```{r write_Seurat}
write_h5ad(sce_obj, tempfile(fileext = ".h5ad"))
```

This is equivalent to converting the `SingleCellExperiment` object to an `AnnData` object and then writing it out:

```{r convert_and_write}
adata <- as_AnnData(sce_obj)
write_h5ad(adata, tempfile(fileext = ".h5ad"))
```

You can again customize the conversion process by providing specific mappings for each slot in the `AnnData` object. For more details, see `?as_AnnData`.

Here's an example:

```{r customize_anndata_conversion}
as_AnnData(
  sce_obj,
  x_mapping = "counts",
  layers_mapping = c("csc_counts"),
  obs_mapping = c(metadata1 = "Int", metadata2 = "IntNA"),
  var_mapping = FALSE,
  obsm_mapping = list(X_pca = "X_pca", X_umap = "X_umap"),
  obsp_mapping = TRUE,
  uns_mapping = c("Bool", "IntScalar")
)
```

The mapping arguments can also be passed directly to `write_h5ad()`.

## Session info

```{r sessioninfo}
sessionInfo()
```
