---
title: Software design
package: anndataR
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Software design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
```

# Introduction

`r Biocpkg("anndataR")` is designed to be an R native implementation of the `AnnData` object, inspired by the of the following packages:

* [_anndata_](https://anndata.readthedocs.io/en/stable/) ([scverse/anndata](https://github.com/scverse/anndata)): The Python `anndata` package and on-disk specification
* `r Biocpkg("zellkonverter")` ([theislab/zellkonverter](https://github.com/theislab/zellkonverter)): Convert `AnnData` files to/from `SingleCellExperiment` objects
* `r Githubpkg("mtmorgan/h5ad")`: Read/write `*.h5ad` files natively using `r Biocpkg("rhdf5")`
* `r CRANpkg("anndataR")` ([dynverse/anndata](https://github.com/dynverse/anndata)): An R implementation of the `AnnData` data structures, uses `r CRANpkg("reticulate")` to read/write `*.h5ad` files

Ideally, this package will be a complete replacement for all of these packages, and will be the go-to package for working with `AnnData` files in R.

# Core features

* An `r CRANpkg("R6")` `AnnData` class to work with objects in R
* In-memory (`InMemoryAnnData`), HDF5-backed (`HDF5AnnData`) and Python-backed (`ReticulateAnnData`) back ends with a consistent interface
* Read/write `.h5ad` files natively
* Convert to/from `SingleCellExperiment` objects
* Convert to/from `Seurat` objects

## Planned features

* Zarr-backed back end (`ZarrAnnData`)

# Class diagram

This diagram shows the main `R6` classes provided by the package:

![](diagrams/class_diagram.svg)

**Notation:**

  - `X: Matrix` - variable `X` is of type `Matrix`
  - `*X: Matrix` - variable `X` is abstract (and implemented by subclasses)
  - `as_SingleCellExperiment(): SingleCellExperiment` - function `as_SingleCellExperiment` returns object of type `SingleCellExperiment`
  - `*as_SingleCellExperiment()` - function `as_SingleCellExperiment` is abstract (and implemented by subclasses)
  - Dashed borders indicate classes that are planned but not yet implemented

# Session info

```{r sessioninfo}
sessionInfo()
```
